<link rel="stylesheet" href="html/scianimator.css">
<script src="html/jquery.min.js"></script>
<script src="html/jquery.scianimator.min.js"></script>

```{r setup, include=FALSE}
require(knitr)
opts_chunk$set(
  cache.path = "target/cache/",
  fig.path = "target/figures/",
  message = FALSE, tidy = FALSE
)
opts_knit$set(animation.fun = hook_scianimator)
source("gtfs.r")
```

Sydney new public transport timetable
====================================

Transport Data Exchange (TDX) program
-------------------------------------
You can download current timetable (and traffic) from [TDX](https://tdx.131500.com.au/). Registration is required.

General Transit Feed Specification (GTFS)
-----------------------------------------
GTFS contains several files saved as comma-delimited text and properly escaped i.e. [CSV](http://tools.ietf.org/html/rfc4180). File encoding should be UTF-8 and BOM is acceptable. The minimal feed consist of 6 files:

File | Defines
-----|--------
agency.txt | One or more transit agencies that provide the data in this feed
stops.txt | Individual locations where vehicles pick up or drop off passengers
routes.txt | Transit routes. A route is a group of trips that are displayed to riders as a single service
trips.txt | Trips for each route. A trip is a sequence of two or more stops that occurs at specific time
stop_times.txt | Times that a vehicle arrives at and departs from individual stops for each trip
calendar.txt | Dates for service IDs using a weekly schedule. Specify when service starts and ends, as well as days of the week where service is available

### GTFS dataset structure

```{r gtfs-dot, engine = "dot", fig.cap = "GTFS", cache=TRUE}
// see http://yihui.name/knitr/demo/engines/
graph gtfs {
  node [shape = box]
  subgraph cluster_1 {
    label = "Transport Data Exchange (TDX) program"
    agency -- routes [label = "agency_id", constraint=false]
    routes -- trips [label = "route_id"]
    trips -- stop_times [label = "trip_id"]
    stop_times -- stops [label = "stop_id"]
    shapes -- trips [label = "shape_id"]
    calendar -- trips [label = "service_id"]
    calendar_dates -- trips [label = "service_id"]
  }
  stops -- transfers [label = "stop_id"]
  fare_attributes -- fare_rules [label = "fare_id"]
  fare_rules -- routes [label = "route_id"]
}
```

General Transit Feed Specification (GTFS) is documented at [Google Developers web site](https://developers.google.com/transit/gtfs/reference).

### BOM (byte-order-mark)
If your imported data set has the first column beginning with 'X.' then the input file contains BOM. Fortunately, this was fixed in R 3.0:

> readLines() and scan() (and hence read.table()) in a UTF-8 locale now discard a UTF-8 byte-order-mark (BOM). Such BOMs are allowed but not recommended by the Unicode Standard: however Microsoft applications can produce them and so they are sometimes found on websites.The encoding name "UTF-8-BOM" for a connection will ensure that a UTF-8 BOM is discarded.

```{r bom-example}
make.names("\357\273\277stop_id")
```

In older versions of R you need to rename the first column or remove BOM manually 

Extent of Greater Sydney public transport area
----------------------------------------------
GTFS dataset must have **stops.txt** file which contains stop location (WGS84), name and several other attributes. Let's read the previous timetable:
```{r load-old-timetable, cache=TRUE }
old <- gtfs.load("data/Oct2012")
```

```{r show-old-timetable, results='asis', echo=FALSE }
kable(head(old[["stops"]]), "html")
```

and the new timetable:

```{r load-new-timetable, cache=TRUE }
new <- gtfs.load("data/Oct2013")
```

```{r show-new-timetable, results='asis', echo=FALSE }
kable(head(new[["stops"]]), "html")
```

The **sp** package has **gConvexHull** function which returns the smallest convex polygon that contains all subgeometries, in our case all transit stops.

```{r compute-bbox}
polygons <- rbind(
  data.frame(gtfs.bbox(new, "convex"), type = "new"),
  data.frame(gtfs.bbox(old, "convex"), type = "old")
)
```
Apparently, public transport coverage remains unchanged under the new time table.
```{r, show-bbox, fig.height=10, fig.width=10, cache=TRUE }
# get base map
base <- with(polygons, get_map(paste(mean(y), mean(x), sep = " "), zoom = 7, color = "bw"))

bbox <- coord_map(
  xlim = extendrange(polygons$x), 
  ylim = extendrange(polygons$y)
)

ggmap(base) + 
  geom_path(data = polygons, aes(x, y, colour = type), size = 1) +
  scale_color_manual(values = c("blue", "red")) +
  geom_point(data = new[["stops"]], aes(x = stop_lon, y = stop_lat), size = 0.5, colour = "black") + # transit stops in black
  bbox +
  ggtitle("Greater Sydney public transport area")
```

Basic network statistics
------------------------
Transit Capacity and Quality of Service Manual [TCQSM, 3rd edition](http://www.trb.org/Main/Blurbs/169437.aspx) provides instructions on how to quantitatively review everything from the person-capacity of a rail transit line to the quality of a bus stop. Most of the calculations are complex and in many cases external datasets (e.g. census data) are required to perform those calculations. 

GTFS dataset provided by TDX programme is rather a shapshot than complete timetable. Hence, only single workday from each timetable (old and new) is analysed. The following days are used:
* old timetable: 2013-10-09
* new timetable: 2013-11-06

```{r trips-summary, cache=TRUE }
old.singleday <- gtfs.trips(old, as.POSIXct("2013-10-09"))
new.singleday <- gtfs.trips(new, as.POSIXct("2013-11-06"))

latlon <- old[["stops"]][, c("stop_id", "stop_lat", "stop_lon")]

# add location info to transit stops
old.singleday <- merge(old.singleday, latlon)
old.singleday$time <- with(old.singleday,
  truncDiffTime(min(arrival_time, na.rm  = T), arrival_time) %% 86400
)
old.singleday <- old.singleday[!is.na(old.singleday$arrival_time), ]

new.singleday <- merge(new.singleday, latlon)
new.singleday$time <- with(new.singleday,
  truncDiffTime(min(arrival_time, na.rm = T), arrival_time) %% 86400
)
new.singleday <- new.singleday[!is.na(new.singleday$arrival_time), ]

# http://www.csgnetwork.com/degreelenllavcalc.html
# Selected Specific Latitude = -33.8230854940154
# 1deg lat -> 110 919.17m
# 1deg lon ->  92 575.84m
# this give us 4km hex bins
lat.bin.width <- 4000 / 110919.17
lon.bin.width <- 4000 / 92575.84

xbnds <- c(
  floor(min(polygons$x) / lon.bin.width) * lon.bin.width, 
  ceiling(max(polygons$x) / lon.bin.width) * lon.bin.width
)
ybnds <- c(
  floor(min(polygons$y) / lat.bin.width) * lat.bin.width,
  ceiling(max(polygons$y) / lat.bin.width) * lat.bin.width
)

registerDoMC() # 4 cores, takes ~ 2 x 20 minutes
old.bins <- compute.hex.bins(old.singleday, xbnds, ybnds, lat.bin.width, lon.bin.width)
new.bins <- compute.hex.bins(new.singleday, xbnds, ybnds, lat.bin.width, lon.bin.width)

new.old.bins <- with(
  merge(old.bins, new.bins, by = c("ID", "time"), all = T, suffixes = c(".old", ".new")),
  data.frame(
    ID = ID,
    time = time,
    count.old = ifelse(is.na(count.old), 0, count.old),
    count.new = ifelse(is.na(count.new), 0, count.new),
    count.diff = ifelse(is.na(count.new), 0, count.new) - ifelse(is.na(count.old), 0, count.old),
    lat = ifelse(is.na(y.old), y.new, y.old),
    lon = ifelse(is.na(x.old), x.new, x.old),
    change.old.new = interaction(is.na(y.old), is.na(y.new))
  )
)

```

Sydney transport network has three modes: bus, train, and ferry.

```{r basic-stats, cache=TRUE, echo=FALSE}

old.summary <- gtfs.trips.summary(old.singleday)
new.summary <- gtfs.trips.summary(new.singleday)

old.stats <- timetable.stats(old.summary, old)
new.stats <- timetable.stats(new.summary, new)

```

                | Old timetable | New timetable
----------------|-----------------------------------------:|---------------------------------------:
Number of trips | `r old.stats[["total.trips"]]` | `r new.stats[["total.trips"]]`
 - Bus          | `r old.stats[["bus.trips"]]`   | `r new.stats[["bus.trips"]]`
 - Train        | `r old.stats[["train.trips"]]` | `r new.stats[["train.trips"]]`
 - Ferry        | `r old.stats[["ferry.trips"]]` | `r new.stats[["ferry.trips"]]`
Longest trip    | |
 - Bus          | `r old.stats[["bus.longest"]]`   | `r new.stats[["bus.longest"]]`
 - Train        | `r old.stats[["train.longest"]]` | `r new.stats[["train.longest"]]`
 - Ferry        | `r old.stats[["ferry.longest"]]` | `r new.stats[["ferry.longest"]]`

```{r speed-by-mode, cahce=TRUE, fig.height=10, fig.width=10, fig.show='animate', echo=FALSE }
ggplot(data = old.summary, aes(y = travel_time, x = distance)) + 
  geom_point(aes(colour = route_type), alpha = 0.7) +
  geom_abline(intercept = 0, slope = 3600/31, linetype = "dotted") + # AM peak
  geom_abline(intercept = 0, slope = 3600/42, linetype = "dotted") + # PM peak
  scale_y_continuous(breaks = seq(0, 5, 1) * 3600, labels = seq(0, 5)) +
  xlim(0, 250) + 
  facet_grid(route_type ~ .) +
  ylab("Travel time [hours]") +
  xlab("Distance [km]") +
  ggtitle("Old timetable\n(cars: AM peak: 31km/h, PM peak: 42km/h)")

ggplot(data = new.summary, aes(y = travel_time, x = distance)) + 
  geom_point(aes(colour = route_type), alpha = 0.7) +
  geom_abline(intercept = 0, slope = 3600/31, linetype = "dotted") +
  geom_abline(intercept = 0, slope = 3600/42, linetype = "dotted") +
  scale_y_continuous(breaks = seq(0, 5, 1) * 3600, labels = seq(0, 5)) +
  xlim(0, 250) + 
  facet_grid(route_type ~ .) +
  ylab("Travel time [hours]") +
  xlab("Distance [km]") +
  ggtitle("New timetable\n(cars: AM peak: 31km/h, PM peak: 42km/h)")
```

Public transport availability during a workday
----------------------------------------------

```{r, network-changes-animation, fig.height=10, fig.width=10, fig.show='animate', results='hide'}
opts_chunk$set(animation.fun = hook_scianimator)
values <- unique(c(seq(min(new.old.bins$count.diff), 0, length.out = 6), seq(0, max(new.old.bins$count.diff), length.out = 6)))
mid = abs(min(new.old.bins$count.diff) / max(new.old.bins$count.diff))
rescaled <- unique(c(
  rescale(seq(min(new.old.bins$count.diff), 0, length.out = 6), c(0, mid)), 
  rescale(seq(0, max(new.old.bins$count.diff), length.out = 6), c(mid, 1))
))
pal <- brewer.pal(11, "RdYlBu")
pal[5] <- brewer.pal(9, "Greens")[4] 
# display.brewer.all()
g <- scale_fill_gradientn(
  limits = range(values),
  colours = pal,
  values = rescaled,
  breaks = round(values, -2),
  labels = round(values, -2),
  name="Difference in available services\n(reduced .. same .. increased)"
)
anim <- llply(sort(unique(new.old.bins$time)), function(t, d, grad) {
  ggmap(base) + 
    geom_hex(data = d[d$time == t, ], stat="identity", aes(x = lon, y = lat, fill = count.diff), alpha = 0.8) +
    grad + bbox +
    ggtitle(format(trunc(Sys.time(), "day") + t, "Public transport availability at %H:%M")) +
    theme(legend.position="bottom")
}, d = new.old.bins, grad = g)
print(anim)
```

Headway time
------------
One of the important metrics in the transport industry headway time. Headway is a measurement of the distance or time between vehicles in a transit system. It describes distance between subsequent vehicles on the same route.

```{r headway-time, cache=TRUE, echo=FALSE}
old.time <- compute.headway(old.singleday, old[["trips"]])
new.time <- compute.headway(new.singleday, new[["trips"]])

old.time <- merge(old.time, old[["stops"]][, c("stop_id", "stop_lat", "stop_lon")])
new.time <- merge(new.time, new[["stops"]][, c("stop_id", "stop_lat", "stop_lon")])

zz <- merge(old.time, new.time, by = c("stop_id"), suffixes = c(".old", ".new"), all = T)
zz <- zz[complete.cases(zz), ]

# > 0 worse; = 0 same; < 0 improvement
zz$diff <- zz$p50.new - zz$p50.old

ggmap(base) + 
  geom_point(data = zz[zz$diff != 0, ], aes(x = stop_lon.old, y = stop_lat.old, colour = cut(diff, cuts))) +
  bbox

g <- scale_fill_gradient2(
  name="Difference in available services\n(reduced .. same .. increased)"
)
cuts <- 60 * c(0, 1, 5, 10, 15, 30, 60, 6*60, 12*60, 24*60)
cuts <- unique(c(- cuts, cuts))
pal <- c(
  rev(brewer.pal(floor(length(cuts) / 2), "Reds")),
  "#FFFFFF",
  brewer.pal(floor(length(cuts) / 2), "Greens")
)
ggplot(data = zz[zz$diff != 0 & abs(zz$diff < 60000), ], aes(x = stop_lon.old, y = stop_lat.old)) + 
  geom_point(aes(colour = diff), size = 0.8) +
  bbox + scale_color_gradientn(colours = pal) + 
  ggtitle("Headway time change") +
  theme(legend.position="bottom")

```

Notes
-----

Sources
-------
[Annual Speed and Traffic Volume Data in Sydney](http://www.rms.nsw.gov.au/publicationsstatisticsforms/downloads/annual_speed_and_traffic_volume_data_2009-2010.pdf)
