```{r setup, include=FALSE}
require(knitr)
opts_chunk$set(
  cache.path = "target/cache/",
  fig.path = "target/figures/",
  message = FALSE, tidy = FALSE
)
opts_knit$set(animation.fun = hook_scianimator)
source("gtfs.r")
```

Sydney new public transport timetable
====================================

Transport Data Exchange (TDX) program
-------------------------------------
You can download current timetable (and traffic) from [TDX](https://tdx.131500.com.au/). Registration is required.

General Transit Feed Specification (GTFS)
-----------------------------------------
GTFS contains several files saved as comma-delimited text and properly escaped i.e. [CSV](http://tools.ietf.org/html/rfc4180). File encoding should be UTF-8 and BOM is acceptable. The minimal feed consist of 6 files:

File | Defines
-----|--------
agency.txt | One or more transit agencies that provide the data in this feed
stops.txt | Individual locations where vehicles pick up or drop off passengers
routes.txt | Transit routes. A route is a group of trips that are displayed to riders as a single service
trips.txt | Trips for each route. A trip is a sequence of two or more stops that occurs at specific time
stop_times.txt | Times that a vehicle arrives at and departs from individual stops for each trip
calendar.txt | Dates for service IDs using a weekly schedule. Specify when service starts and ends, as well as days of the week where service is available

```{r gtfs-dot, engine = "dot", fig.cap = "GTFS", cache=TRUE}
// see http://yihui.name/knitr/demo/engines/
graph gtfs {
  node [shape = box]
  subgraph cluster_1 {
    label = "Transport Data Exchange (TDX) program"
    agency -- routes [label = "agency_id", constraint=false]
    routes -- trips [label = "route_id"]
    trips -- stop_times [label = "trip_id"]
    stop_times -- stops [label = "stop_id"]
    shapes -- trips [label = "shape_id"]
    calendar -- trips [label = "service_id"]
    calendar_dates -- trips [label = "service_id"]
  }
  stops -- transfers [label = "stop_id"]
  fare_attributes -- fare_rules [label = "fare_id"]
  fare_rules -- routes [label = "route_id"]
}
```

General Transit Feed Specification (GTFS) is documented at [Google Developers web site](https://developers.google.com/transit/gtfs/reference).

### BOM (byte-order-mark)
If your imported data set has the first column beginning with 'X.' then the input file contains BOM. Fortunately, this was fixed in R 3.0:

> readLines() and scan() (and hence read.table()) in a UTF-8 locale now discard a UTF-8 byte-order-mark (BOM). Such BOMs are allowed but not recommended by the Unicode Standard: however Microsoft applications can produce them and so they are sometimes found on websites.The encoding name "UTF-8-BOM" for a connection will ensure that a UTF-8 BOM is discarded.

```{r bom-example}
make.names("\357\273\277stop_id")
```

In older versions of R you need to rename the first column or remove BOM manually 

Extent of Greater Sydney public transport area
----------------------------------------------
GTFS dataset must have **stops.txt** file which contains stop location (WGS84), name and several other attributes. Let's read the previous timetable:
```{r load-old-timetable, cache=TRUE }
old <- gtfs.load("data/Oct2012")
```

```{r show-old-timetable, results='asis', echo=FALSE }
kable(head(old[["stops"]]), "html")
```

and the new timetable:

```{r load-new-timetable, cache=TRUE }
new <- gtfs.load("data/Oct2013")
```

```{r show-new-timetable, results='asis', echo=FALSE }
kable(head(new[["stops"]]), "html")
```

The **sp** package has **gConvexHull** function which returns the smallest convex polygon that contains all subgeometries, in our case all transit stops.

```{r compute-bbox}
polygons <- rbind(
  data.frame(gtfs.bbox(new, "convex"), type = "new"),
  data.frame(gtfs.bbox(old, "convex"), type = "old")
)
```
Apparently, public transport coverage remains unchanged under the new time table.
```{r, show-bbox, fig.height=10, fig.width=10, cache=TRUE }
# get base map
base <- with(polygons, get_map(paste(mean(y), mean(x), sep = " "), zoom = 7, color = "bw"))

bbox <- coord_map(
  xlim = extendrange(polygons$x), 
  ylim = extendrange(polygons$y)
)

ggmap(base) + 
  geom_path(data = polygons, aes(x, y, colour = type), size = 1) +
  scale_color_manual(values = c("blue", "red")) +
  geom_point(data = new[["stops"]], aes(x = stop_lon, y = stop_lat), size = 0.5, colour = "black") + # transit stops in black
  bbox +
  ggtitle("Greater Sydney public transport area")
```

Basic network statistics
------------------------
Sydney transport network has three modes: bus, train, and ferry.

```{r trips-summary, cache=TRUE }
old.singleday <- gtfs.trips(old, as.POSIXct("2013-10-09"))
new.singleday <- gtfs.trips(new, as.POSIXct("2013-11-06"))

latlon <- old[["stops"]][, c("stop_id", "stop_lat", "stop_lon")]

# add location info to transit stops
old.singleday <- merge(old.singleday[, c("trip_id", "stop_id", "arrival_time", "route_type")], latlon)
old.singleday$time <- with(old.singleday,
  truncDiffTime(min(arrival_time, na.rm  = T), arrival_time) %% 86400
)
old.singleday <- old.singleday[!is.na(old.singleday$arrival_time), ]

new.singleday <- merge(new.singleday[, c("trip_id", "stop_id", "arrival_time", "route_type")], latlon)
new.singleday$time <- with(new.singleday,
  truncDiffTime(min(arrival_time, na.rm = T), arrival_time) %% 86400
)
new.singleday <- new.singleday[!is.na(new.singleday$arrival_time), ]

# http://www.csgnetwork.com/degreelenllavcalc.html
# Selected Specific Latitude = -33.8230854940154
# 1deg lat -> 110 919.17m
# 1deg lon ->  92 575.84m
# this give us 4km hex bins
lat.bin.width <- 4000 / 110919.17
lon.bin.width <- 4000 / 92575.84

xbnds <- c(
  floor(min(polygons$x) / lon.bin.width) * lon.bin.width, 
  ceiling(max(polygons$x) / lon.bin.width) * lon.bin.width
)
ybnds <- c(
  floor(min(polygons$y) / lat.bin.width) * lat.bin.width,
  ceiling(max(polygons$y) / lat.bin.width) * lat.bin.width
)

registerDoMC() # 4 cores, takes ~ 2 x 20 minutes
old.bins <- compute.hex.bins(old.singleday, xbnds, ybnds, lat.bin.width, lon.bin.width)
new.bins <- compute.hex.bins(new.singleday, xbnds, ybnds, lat.bin.width, lon.bin.width)

new.old.bins <- with(
  merge(old.bins, new.bins, by = c("ID", "time"), all = T, suffixes = c(".old", ".new")),
  data.frame(
    ID = ID,
    time = time,
    count.old = ifelse(is.na(count.old), 0, count.old),
    count.new = ifelse(is.na(count.new), 0, count.new),
    count.diff = ifelse(is.na(count.new), 0, count.new) - ifelse(is.na(count.old), 0, count.old),
    lat = ifelse(is.na(y.old), y.new, y.old),
    lon = ifelse(is.na(x.old), x.new, x.old),
    change.old.new = interaction(is.na(y.old), is.na(y.new))
  )
)

```

Some animation
```{r, network-changes-animation, fig.show='animate'}
opts_chunk$set(animation.fun = hook_scianimator)
values <- unique(c(seq(min(new.old.bins$count.diff), 0, length.out = 6), seq(0, max(new.old.bins$count.diff), length.out = 6)))
mid = abs(min(new.old.bins$count.diff) / max(new.old.bins$count.diff))
rescaled <- unique(c(
  rescale(seq(min(new.old.bins$count.diff), 0, length.out = 6), c(0, mid)), 
  rescale(seq(0, max(new.old.bins$count.diff), length.out = 6), c(mid, 1))
))
# display.brewer.all()
g <- scale_fill_gradientn(
  limits = range(values),
  colours = brewer.pal(11, "RdYlBu"),
  values = rescaled,
  breaks = round(values, -2),
  labels = round(values, -2),
  name="Difference in available services\n(reduced .. same .. increased)"
)
l_ply(sort(unique(new.old.bins$time)), function(t, d, grad) {
  g <- ggmap(base) + 
    geom_hex(data = d[d$time == t, ], stat="identity", aes(x = lon, y = lat, fill = count.diff), alpha = 0.8) +
    grad +
    ggtitle(format(trunc(Sys.time(), "day") + t, "%H:%M")) +
    theme(legend.position="bottom")
  print(g)
}, d = new.old.bins, grad = g)
```

