```{r setup, include=FALSE}
require(knitr)
opts_chunk$set(cache=TRUE, message=FALSE, tidy=FALSE)

source("src/main/R/gtfs.r")
```

Sydney new public transport timetable
====================================

Transport Data Exchange (TDX) program
-------------------------------------
You can download current timetable (and traffic) from [TDX](https://tdx.131500.com.au/). Registration is required.

General Transit Feed Specification (GTFS)
-----------------------------------------
GTFS contains several files saved as comma-delimited text and properly escaped i.e. [CSV](http://tools.ietf.org/html/rfc4180). File encoding should be UTF-8 and BOM is acceptable. The minimal feed consist of 6 files:

File | Defines
-----|--------
agency.txt | One or more transit agencies that provide the data in this feed
stops.txt | Individual locations where vehicles pick up or drop off passengers
routes.txt | Transit routes. A route is a group of trips that are displayed to riders as a single service
trips.txt | Trips for each route. A trip is a sequence of two or more stops that occurs at specific time
stop_times.txt | Times that a vehicle arrives at and departs from individual stops for each trip
calendar.txt | Dates for service IDs using a weekly schedule. Specify when service starts and ends, as well as days of the week where service is available

General Transit Feed Specification (GTFS) is documented at [Google Developers web site](https://developers.google.com/transit/gtfs/reference).

### BOM (byte-order-mark)
If your imported data set has the first column beginning with 'X.' then the input file contains BOM. Fortunately, this was fixed in R 3.0:

> readLines() and scan() (and hence read.table()) in a UTF-8 locale now discard a UTF-8 byte-order-mark (BOM). Such BOMs are allowed but not recommended by the Unicode Standard: however Microsoft applications can produce them and so they are sometimes found on websites.The encoding name "UTF-8-BOM" for a connection will ensure that a UTF-8 BOM is discarded.

```{r}
make.names("\357\273\277stop_id")
```

In older versions of R you need to rename the first column or remove BOM manually 

Extent of Greater Sydney public transport area
----------------------------------------------
GTFS dataset must have **stops.txt** file which contains stop location (WGS84), name and several other attributes. Let's read the previous timetable:
```{r}
old <- gtfs.load("data/Oct2012")
```

```{r results='asis', echo=FALSE }
kable(head(old[["stops"]]), "html")
```

and the new timetable:

```{r}
new <- gtfs.load("data/Oct2013")
```

```{r results='asis', echo=FALSE }
kable(head(new[["stops"]]), "html")
```

The **sp** package has **gConvexHull** function which returns the smallest convex polygon that contains all subgeometries, in our case all transit stops.

```{r}
polygons <- rbind(
  data.frame(gtfs.bbox(new, "convex"), type = "new"),
  data.frame(gtfs.bbox(old, "convex"), type = "old")
)
```
Apparently, public transport coverage remains unchanged under the new time table.
```{r, fig.height=10, fig.width=10 }
# get base map
base <- with(polygons, get_map(paste(mean(y), mean(x), sep = " "), zoom = 7, color = "bw"))

bbox <- coord_map(
  xlim = extendrange(polygons$x), 
  ylim = extendrange(polygons$y)
)

ggmap(base) + 
  geom_path(data = polygons, aes(x, y, colour = type), size = 1) +
  scale_color_manual(values = c("blue", "red")) +
  geom_point(data = new[["stops"]], aes(x = stop_lon, y = stop_lat), size = 0.5, colour = "black") + # transit stops in black
  bbox +
  ggtitle("Greater Sydney public transport area")
```


